A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization

### 2020-10-18 ms/ms.tex
pre-release

  \textsuperscript{1} School of Life Sciences, University of Hawaii,  Honolulu,  Hawaii,  USA
  }

@@

Stomatal pores control both leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and defense. Anatomical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes could permanently reduce the gas exchange capacity for the life of the leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which partially determine maximum rates of gas exchange. The model predicts greater stomatal size or density increases the probability of colonization, but the effect is most pronounced when the fraction of leaf surface covered by stomata is low. I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential anatomical conflict between limiting pathogen colonization and minimizing the fraction of leaf surface covered by stomata. Although a connection between gas exchange and pathogen defense has been suggested empirically, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. A limitation of the model is that it does not include variation in innate immunity and stomatal closure in response to pathogens. Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for precise predictions.

@@

The stomatal density and maximum pore area set an anatomical upper limit
on stomatal conductance (Brown and Escombe, 1900; Parlange and Waggoner,
1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b; Lehmann and
Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019), but stomatal
shape, distribution, and patterning also affect gas exchange. Smaller
guard cells and dumbbell-shaped stomata of grasses can respond faster to
environmental changes (Drake et al., 2013), but responsiveness is
further modulated by subsidiary cell anatomy and physiology (Franks and
Farquhar, 2007; Raissig et al., 2017; Gray et al., 2020). Stomatal
clustering reduces gas exchange and photosynthesis because adjacent
stomata interfere with one another (Dow et al., 2014b), diffusion shells
overlap (Lehmann and Or, 2015), and limitations on lateral diffusion of
CO\(_2\) in the mesophyll (Lawson and Blatt, 2014 and references
therein). However, sparse clusters of small stomata could allow a leaf
with low rates of gas exchange to have faster stomatal response compared
to a leaf with large, low-density stomata (Papanatsiou et al., 2017).
Leaves with stomata on both lower and upper surfaces (amphistomatic or
amphistomatous) supply more CO\(_2\) to the mesophyll than hypostomatic
(aka hypostomatous) leaves that only have stomata on the lower surface
(Parkhurst, 1978; Gutschick, 1984; Parkhurst and Mott, 1990; Oguchi et
al., 2018). In addition to anatomy, the pore area shrinks and expands in
response to internal and external factors to regulate gas exchange
dynamically (Buckley, 2019). For example, stomata typically open during
the day and close at night in C\(_3\)/C\(_4\) plants, but the opposite
is true for CAM plants. Shade, high vapor pressure deficits, dry soil
and other factors can cause stomata to (partially) close even in the
middle of the day. Variation in how stomata respond to internal and
external signals may explain as much of the variation in gas exchange
across leaves as anatomy (Lawson and Blatt, 2014).

@@

Stomatal anatomy could be a key link between gas exchange and pathogen
colonization. Although many anatomical factors and stomatal movement
affect gas exchange (see above), here I focus on the density and size of
stomata in a hypostomatous leaf. Stomatal size refers to both the area
of guard cells when fully open, from which one can calculate the pore
area for gas exchange (see \protect\hyperlink{model}{Model}). For
simplicity, I model a hypostomatous leaf, but consider the implications
for amphistomatous leaves in the
\protect\hyperlink{discussion}{Discussion}. Stomatal size and density
not only determine the theoretical maximum stomatal conductance
(\gsmax), but is also proportional to the operational stomatal
conductance (\gsop) in many circumstances (Franks et al., 2009, 2014;
Dow et al., 2014a; McElwain et al., 2016; Murray et al., 2019). \gsop{}
is the actual stomatal conductance of plants in the field and is almost
always below \gsmax{} because stomata are usually not fully open.
Although they are not the same, the strong empirical relationship
between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used
as a proxy for \gsop{} without explicitly modeling dynamic changes in
stomatal aperture (see \protect\hyperlink{discussion}{Discussion}).
Stomatal size and density have also been measured on many more species
than stomatal responsiveness, which may make it easier to test
predictions.

@@

In this section, I introduce a spatially explicit model of successful
pathogen colonization on a leaf surface. I explain the model structure
and assumptions here; the
\protect\hyperlink{materials-and-methods}{Materials and Methods} section
below describes how I analyzed the model to address the goals of the
study. For generality, I refer to a generic ``pathogen'' that lands on
leaf and moves to a stomate. The model is agnostic to the type of
pathogen (virus, bacterium, fungus, etc.) and the specific biological
details of how it moves. For example, motile bacterial cells can land
and move around (Beattie and Lindow, 1995) whereas fungi may germinate
from a cyst and grow until they form an appresorium for infection
(Tucker and Talbot, 2001). These very different tropic movements on the
leaf are treated identically here. I do not model photosynthesis
explicitly, but assume that stomatal conductance limits carbon fixation,
even though the relationship is nonlinear. I used Sympy version 1.6.1
(Meurer et al., 2017) for symbolic derivations.

@@

Stomata develop relatively equal spacing to minimize resistance to
lateral diffusion (Morison et al., 2005), allow space between stomata
(Dow et al., 2014b), and prevent stomatal interference (Lehmann and Or,
2015). Here I assume that stomata are arrayed in an equilateral
triangular grid with a density \(D\) and size (area) \(S\) on the
abaxial surface only, since most leaves are hypostomatoues {[}Muir
(2015); but see \protect\hyperlink{discussion}{Discussion}{]}. This
assumption ignores veins, trichomes, and within-leaf variation in
stomatal density. Stomata are therefore arrayed in an evenly spaced grid
(Figure \ref{fig:fig1}a). The interstomatal distance \(U\), measured as
the distance from the center of one stomata to the next, is the maximal
diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
boundary between neighboring stomata. The area of a hexagon is
\(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
stomatal density is the inverse of this area, such that
\(D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}\).
Therefore, interstomatal distance can be derived from the stomatal
density as:

@@

The \protect\hyperlink{model}{Model} calculates a probability of host
colonization (Equation \ref{eq:p_colonize}) as a function of stomatal
density, size, and position of a pathogen on the leaf. I solved
\(p_\text{colonize}\) by importing symbolic derivations from Sympy into
R with \textbf{reticulate} version 1.16 (Ushey et al., 2020) and used
the \texttt{integral2()} function in the \textbf{pracma} package version
2.2.9 (Borchers, 2019) for numerical integration. I used R version 4.0.2
(R Core Team, 2020) for all analyses and wrote the paper in
\textbf{rmarkdown} version 2.3 (Xie et al., 2018; Allaire et al., 2020).
Citations for additional R software packages are in Appendix 2. Source
code is deposited on GitHub
(\url{https://github.com/cdmuir/stomata-tradeoff}) and will be archived
on Zenodo upon publication.

@@

The probability of colonization (\(p_\text{colonize}\)) is not simply
proportional to stomatal cover (\fs). At low \fs, \(p_\text{colonize}\)
increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a). At
higher \fs, \(p_\text{colonize}\) increases linearly with \fs. When
pathogens persist indefinitely (\(H = 0\)), any combination of stomatal
size (\(S\)) and density (\(D\)) with the same \fs{} have the same
effect on \(p_\text{colonize}\). When \(H > 0\), pathogens are less
likely to land close enough to a stomate to infect before dying, so
\(p_\text{colonize}\) is closer to \fs{} (Figure \ref{fig:fig3}a). The
maximum \(p_\text{colonize}\) under the range of parameters considered
was \(\sim 0.6\) when \(H = 0\) and \fs{} is at its maximum value of
1/3. When \fs{} is low, \(p_\text{colonize}\) is also low. The
relationship between \(p_\text{colonize}\), \fs, and \gsmax is
qualitatively similar in the spatially implicit model, but the values
for \(p_\text{colonize}\) are substantially higher because pathogens can
potentially colonize any stomate on the leaf rather than only those in
the focal triangle (see
\protect\hyperlink{appendix-1-spatially-implicit-model}{Appendix 1:
Spatially implicit model} for more detail). Bear in mind that this is
the probability for a single individual searching randomly; if enough
individuals reach the leaf and/or they can actively find stomata, it's
almost certain that at least some will colonize the leaf. However,
reducing \(p_\text{colonize}\) may help plants limit the damage since
fewer total individual pathogens will colonize the leaf interior.

@@

\begin{longtable}[]{@{}ll@{}}
\toprule
\begin{minipage}[b]{0.50\columnwidth}\raggedright
Model Prediction\strut
\end{minipage} & \begin{minipage}[b]{0.44\columnwidth}\raggedright
How to test it\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.50\columnwidth}\raggedright
Increasing stomatal size and/or density will have a larger effect on
pathogen colonization in leaves with low stomatal cover.\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Compare the effect of changing stomatal size and/or density on pathogen
colonization in leaves with low and high stomatal cover.\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.50\columnwidth}\raggedright
Increasing stomatal size and/or density will have a smaller effect on
pathogen colonization in ephiphytic environments more hostile to
pathogens.\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Compare the effect of changing stomatal size and/or density on pathogen
colonization in more hostile environments (e.g.~drier, higher
light)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.50\columnwidth}\raggedright
When selection against pathogen colonization stronger, the stomatal
size-density scaling exponent should be lower\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Measure stomatal anatomy in environments that differ in pathogen
colonization using comparative or experimental approaches\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

I would like to thank Athena McKown and an anonymous referee for
valuable feedback that improved this manuscript.

@@

I am grateful startup funds from the University of Hawaii for supporting
this work.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-allaire_rmarkdown:_2020}{}%
Allaire, J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Wickham,
H., Cheng, J., Chang, W., and Iannone, R. (2020). Rmarkdown: Dynamic
Documents for R. Available at:
\url{https://github.com/rstudio/rmarkdown}.

@@

\leavevmode\hypertarget{ref-zeng_plant_2010}{}%
Zeng, W., Melotto, M., and He, S. Y. (2010). Plant stomata: A checkpoint
of host immunity and pathogen virulence. \emph{Current Opinion in
Biotechnology} 21, 599--603.
doi:\href{https://doi.org/10.1016/j.copbio.2010.05.006}{10.1016/j.copbio.2010.05.006}.

>>>>

  \textsuperscript{1} School of Life Sciences, University of
Hawai'i,  Honolulu,  Hawaii,  USA
  }

@@

Stomatal pores control leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and pathogen colonization. Anatomical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes could permanently reduce the gas exchange capacity for the life of the leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which partially determine maximum rates of gas exchange. The model predicts greater stomatal size or density increases the probability of colonization, but the effect is most pronounced when the fraction of leaf surface covered by stomata is low. I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential anatomical conflict between limiting pathogen colonization and minimizing the fraction of leaf surface covered by stomata. Although a connection between gas exchange and pathogen defense has been suggested empirically, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. A limitation of the model is that it does not include variation in innate immunity and stomatal closure in response to pathogens. Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal size and density among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for precise predictions.

@@

The stomatal density and maximum pore area set an anatomical upper limit
on stomatal conductance (Brown and Escombe, 1900; Parlange and Waggoner,
1970; Franks and Farquhar, 2001; Franks and Beerling, 2009b; Lehmann and
Or, 2015; Sack and Buckley, 2016; Harrison et al., 2019), but stomatal
shape, distribution, and patterning also affect gas exchange. Smaller
guard cells and dumbbell-shaped stomata of grasses can respond faster to
environmental changes (Drake et al., 2013), but responsiveness is
further modulated by subsidiary cell anatomy and physiology (Franks and
Farquhar, 2007; Raissig et al., 2017; Gray et al., 2020). Stomatal
clustering reduces gas exchange and photosynthesis because adjacent
stomata interfere with one another (Dow et al., 2014b), diffusion shells
overlap (Lehmann and Or, 2015), and limitations on lateral diffusion of
CO\(_2\) in the mesophyll (Lawson and Blatt, 2014 and references
therein). However, sparse clusters of small stomata could allow a leaf
with low rates of gas exchange to have faster stomatal response compared
to a leaf with large, low-density stomata (Papanatsiou et al., 2017).
Leaves with stomata on both lower and upper surfaces (amphistomatous)
supply more CO\(_2\) to the mesophyll than hypostomatous leaves that
only have stomata on the lower surface (Parkhurst, 1978; Gutschick,
1984; Parkhurst and Mott, 1990; Oguchi et al., 2018). In addition to
anatomy, the pore area shrinks and expands in response to internal and
external factors to regulate gas exchange dynamically (Buckley, 2019).
For example, stomata typically open during the day and close at night in
C\(_3\)/C\(_4\) plants, but the opposite is true for CAM plants. Shade,
high vapor pressure deficits, dry soil and other factors can cause
stomata to (partially) close even in the middle of the day. Variation in
how stomata respond to internal and external signals may explain as much
of the variation in gas exchange across leaves as anatomy (Lawson and
Blatt, 2014).

@@

Stomatal anatomy could be a key link between gas exchange and pathogen
colonization. Although many anatomical factors and stomatal movement
affect gas exchange (see above), here I focus on the density and size of
stomata in a hypostomatous leaf. Stomatal size refers to both the area
of guard cells when fully open, from which one can calculate the pore
area for gas exchange (see \protect\hyperlink{model}{Model}). For
simplicity, I model a hypostomatous leaf, but consider the implications
for amphistomatous leaves in the
\protect\hyperlink{discussion}{Discussion}. Stomatal size and density
not only determine the theoretical maximum stomatal conductance
(\gsmax), but are also proportional to the operational stomatal
conductance (\gsop) in many circumstances (Franks et al., 2009, 2014;
Dow et al., 2014a; McElwain et al., 2016; Murray et al., 2019). \gsop{}
is the actual stomatal conductance of plants in the field and is almost
always below \gsmax{} because stomata are usually not fully open.
Although they are not the same, the strong empirical relationship
between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used
as a proxy for \gsop{} without explicitly modeling dynamic changes in
stomatal aperture (see \protect\hyperlink{discussion}{Discussion}).
Stomatal size and density have also been measured on many more species
than stomatal responsiveness, which may make it easier to test
predictions.

@@

For generality, I refer to a generic ``pathogen'\,' that lands on leaf
and moves to a stomate. The model is agnostic to the type of pathogen
(virus, bacterium, fungus, etc.) and the specific biological details of
how it moves. For example, motile bacterial cells can land and move
around (Beattie and Lindow, 1995) whereas fungi may germinate from a
cyst and grow until they form an appresorium for infection (Tucker and
Talbot, 2001). These very different tropic movements on the leaf are
treated identically here. I do not model photosynthesis explicitly, but
assume that stomatal conductance limits carbon fixation, even though the
relationship is nonlinear. I used Sympy version 1.6.1 (Meurer et al.,
2017) for symbolic derivations.

@@

Stomata develop relatively equal spacing to minimize resistance to
lateral diffusion (Morison et al., 2005), allow space between stomata
(Dow et al., 2014b), and prevent stomatal interference (Lehmann and Or,
2015). Here I assume that stomata are arrayed in an equilateral
triangular grid with a density \(D\) and size (area) \(S\) on the
abaxial surface only, since most leaves are hypostomatous {[}Muir
(2015); but see \protect\hyperlink{discussion}{Discussion}{]}. This
assumption ignores veins, trichomes, and within-leaf variation in
stomatal density. Stomata are therefore arrayed in an evenly spaced grid
(Figure \ref{fig:fig1}a). The interstomatal distance \(U\), measured as
the distance from the center of one stomata to the next, is the maximal
diagonal of the hexagon in \(\mu\textrm{m}\) that forms an equal area
boundary between neighboring stomata. The area of a hexagon is
\(A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2\). By definition the
stomatal density is the inverse of this area, such that
\(D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}\).
Therefore, interstomatal distance can be derived from the stomatal
density as:

@@

The \protect\hyperlink{model}{Model} calculates a probability of host
colonization (Equation \ref{eq:p_colonize}) as a function of stomatal
density, size, and position of a pathogen on the leaf. I solved
\(p_\text{colonize}\) by importing symbolic derivations from Sympy into
R with \textbf{reticulate} version 1.16 (Ushey et al., 2020) and used
the \texttt{integral2()} function in the \textbf{pracma} package version
2.2.9 (Borchers, 2019) for numerical integration. I used R version 4.0.2
(R Core Team, 2020) for all analyses and wrote the paper in
\textbf{rmarkdown} version 2.4 (Xie et al., 2018; Allaire et al., 2020).
Citations for additional R software packages are in Appendix 2. Source
code is deposited on GitHub
(\url{https://github.com/cdmuir/stomata-tradeoff}) and will be archived
on Zenodo upon publication.

@@

The probability of colonization (\(p_\text{colonize}\)) is not simply
proportional to stomatal cover (\fs). At low \fs, \(p_\text{colonize}\)
increases rapidly relative to \fs{} at first (Figure \ref{fig:fig3}a).
At higher \fs, \(p_\text{colonize}\) increases linearly with \fs. When
pathogens persist indefinitely (\(H = 0\)), any combination of stomatal
size (\(S\)) and density (\(D\)) with the same \fs{} have the same
effect on \(p_\text{colonize}\). When \(H > 0\), pathogens are less
likely to land close enough to a stomate to infect before dying, so
\(p_\text{colonize}\) is closer to \fs{} (Figure \ref{fig:fig3}a). The
maximum \(p_\text{colonize}\) under the range of parameters considered
was \(\sim 0.6\) when \(H = 0\) and \fs{} is at its maximum value of
1/3. When \fs{} is low, \(p_\text{colonize}\) is also low. The
relationship between \(p_\text{colonize}\), \fs, and \gsmax is
qualitatively similar in the spatially implicit model, but the values
for \(p_\text{colonize}\) are substantially higher because pathogens can
potentially colonize any stomate on the leaf rather than only those in
the focal triangle (see
\protect\hyperlink{appendix-1-spatially-implicit-model}{Appendix 1:
Spatially implicit model} for more detail). Bear in mind that this is
the probability for a single individual searching randomly; if enough
individuals reach the leaf and/or they can actively find stomata, it's
almost certain that at least some will colonize the leaf. However,
reducing \(p_\text{colonize}\) may help plants limit the damage since
fewer total individual pathogens will colonize the leaf interior.

@@

\begin{longtable}[]{@{}ll@{}}
\toprule
\begin{minipage}[b]{0.50\columnwidth}\raggedright
Model Prediction\strut
\end{minipage} & \begin{minipage}[b]{0.44\columnwidth}\raggedright
How to test it\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.50\columnwidth}\raggedright
Increasing stomatal size and/or density will have a larger effect on
pathogen colonization in leaves with low stomatal cover.\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Compare the effect of changing stomatal size and/or density on pathogen
colonization in leaves with low and high stomatal cover.\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.50\columnwidth}\raggedright
Increasing stomatal size and/or density will have a smaller effect on
pathogen colonization in ephiphytic environments more hostile to
pathogens.\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Compare the effect of changing stomatal size and/or density on pathogen
colonization in more hostile environments (e.g.~drier, higher
light)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.50\columnwidth}\raggedright
When selection against pathogen colonization is stronger, the stomatal
size-density scaling exponent should be lower\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright
Measure stomatal anatomy in environments that differ in pathogen
colonization using comparative or experimental approaches\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

I would like to thank Athena McKown and Karl Fetter for valuable
feedback that improved this manuscript.

@@

I am grateful startup funds from the University of Hawai'i for
supporting this work.

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-allaire_rmarkdown:_2020}{}%
Allaire, J., Xie, Y., McPherson, J., Luraschi, J., Ushey, K., Wickham,
H., Cheng, J., Chang, W., and Iannone, R. (2020). Rmarkdown: Dynamic
Documents for R. Available at:
\url{https://github.com/rstudio/rmarkdown}.

@@

\leavevmode\hypertarget{ref-zeng_plant_2010}{}%
Zeng, W., Melotto, M., and He, S. Y. (2010). Plant stomata: A checkpoint
of host immunity and pathogen virulence. \emph{Current Opinion in
Biotechnology} 21, 599--603.
doi:\href{https://doi.org/10.1016/j.copbio.2010.05.006}{10.1016/j.copbio.2010.05.006}.
\end{cslreferences}
